---
title: "Exploring assumptions"
author: "Petri Haavisto"
format: html
editor: visual
---

## How to check the assumptions... Chapter 5

... Before deciding which statistical test is appropriate. If you use a parametric test when your data is not parametric, result are likely to be inaccurate.

Pipe=Ctrl+Shift+M\
Left Arrow = Alt and - (minus)

1 First libraries we are going to use.

```{r}
library(conflicted)  
library(tidyverse)
library(car)
library(ggplot2)
library(pastecs)
library(psych)
library(effects)
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

2 Next we read raw_data.csv into a tibble. Tibble is a modern version of data frame.

```{r}
raw <- here::here("data/grip_test_1.csv") |> readr::read_csv()
view(raw)
```

TEST 1 Let's load festival data as a test data:

```{r}
dlf <- read.delim(file <- here::here("data/DownloadFestival.dat"), header = TRUE)
view(dlf)
```

TEST 2 Histogram of festival data:

```{r}
hist.dayone <- ggplot(dlf, aes(day1)) + theme(legend.position = "none") +
  geom_histogram(aes(y = ..density..), colour = "black", fill = "white", binwidth = 0.1)+
    labs(x = "Hygiene score on day 1", y = "Density") + 
   stat_function(fun = dnorm, args = list(mean = mean(dlf$day1, na.rm = TRUE), sd = sd(dlf$day1, na.rm = TRUE)), colour = "black", size = 1)
hist.dayone
```

TEST 3

```{r}
qqPlot(dlf$day2, ylab = "Your label here", id = FALSE)
```

3 Histogram of dominating hand grip values:

```{r}
hist.dgrip <- ggplot(raw, aes(MAX_HGS_Dom))+
  geom_histogram(aes(y = ..density..), colour = "black", fill = "white", binwidth = 2)+
    labs(x = "Strength of the dominating Hand", y = "Density") + 
   stat_function(fun = dnorm, args = list(mean = mean(raw$MAX_HGS_Dom), sd = sd(raw$MAX_HGS_Dom)), colour = "black", size = 1)
hist.dgrip
```

Q-Q Plot of the Strength of the dominating hand.

```{r}
qqplot.dgrip <- qqPlot(raw$MAX_HGS_Dom, ylab = "MAX_HGS_Dom", id = TRUE)
#qqPlot(raw$Age, ylab = "Age")
```

```{r}
fivenum(raw$MAX_HGS_Dom)
```

```{r}
festival.test <- stat.desc(dlf[, c("day1", "day2", "day3")], basic = FALSE, norm = TRUE)
view(festival.test)
```

```{r}
test <- c(-11,0, 1, 2, 3, 4, 5, 6, 7)
qqplot.dgrip <- qqPlot(test)
test.stats <- stat.desc(test, basic = FALSE, norm = TRUE)
view(test.stats)
```

```{r}
dom.hand <- stat.desc(raw$MAX_HGS_Dom, basic = FALSE, norm = TRUE)
view(dom.hand)
age.stats <- stat.desc(raw$Age, basic = FALSE, norm = TRUE)
view(age.stats)
```

```{r}
qqnorm(raw$MAX_HGS_Dom, pch = 1, frame = FALSE)
qqline(raw$MAX_HGS_Dom, col = "steelblue", lwd = 2)
```

```{r}
qqnorm(raw$Age, pch = 1, frame = FALSE)
qqline(raw$Age, col = "steelblue", lwd = 2)
```
