---
title: "Scatterplot Example 1"
author: "Petri Haavisto"
format: html
editor: visual
---

## Scatter plot Example 1

Hot to use ggplot to create beautiful scatter plots.

Load libraries:

```{r}
library(conflicted)  
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

Let's have look, what data sets **R** has ready for us and this time we are interested of mpg (miles per gallon):

```{r}
data() #gives you list of R data sets
?mpg
names(mpg)
```

Let's try to plot this miles per gallon data. Grammar of graphics: data set, aesthetics, geometry

```{r}
mpg |> 
  filter(hwy < 35) |> #manipulate data befor ggplot
  ggplot(aes(x = displ, #engine displacement
             y = hwy, #highway miles per gallon
             color = drv)) +   #color coding based on variable drv
  geom_point()+
  #smooth lines (inherits color from drv)
  geom_smooth(formula = y ~ x, method = lm, se = F)+ 
  labs(x = "Engine size",
       y = "MPG on the Highway",
       title = "Fuel effiency")+
  theme_minimal()
  #ggsave("MPH_hwy.jpg")
```

Let's plot this again but this time line is not "lm" and we also let "se" (standard error to the plot):

```{r}
mpg |> 
  filter(hwy < 35) |> 
  ggplot(aes(x = displ,
             y = hwy)) +   
  geom_point(aes(color = drv))+
  geom_smooth()+ 
  labs(x = "Engine size",
       y = "MPG on the Highway",
       title = "Fuel effiency")+
  theme_minimal()
  #ggsave("MPH_hwy.jpg")
```

## Test 2 msleep

Let's have look at sleep data

```{r}
data() #gives you list of R data sets
?msleep
view(msleep)
```

Let's try to scatter plot this data: VERSION 1

```{r}
msleep |> 
  filter(bodywt < 2, !is.na(brainwt)) |> 
  ggplot(aes(bodywt, brainwt))+ #these are x and y
  geom_point(aes(color = sleep_total, size = awake))+
  geom_smooth(formula = y ~ x, method = loess)+
  labs(x = "Body Weight", y = "Brain Weight", title = "Brain and body weight")+
  theme_minimal()
```

VERSION 2 (stripped)

```{r}
msleep |> 
  filter(bodywt < 2, !is.na(brainwt)) |> 
  ggplot(aes(bodywt, brainwt))+ #these are x and y
  geom_point()+
  geom_smooth(formula = y ~ x, method = loess)+
  labs(x = "Body Weight", y = "Brain Weight", title = "Brain and body weight")+
  theme_minimal()
```

VERSION 3

```{r}
msleep |> 
  filter(bodywt < 2, !is.na(brainwt)) |> 
  ggplot(aes(bodywt, brainwt))+ #these are x and y
  geom_point(aes(color = sleep_total, size = awake))+
  geom_smooth(formula = y ~ x, method = lm, se = F)+
  labs(x = "Body Weight", y = "Brain Weight", title = "Brain and body weight")+
  theme_minimal()
```
